---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Bootstrapping the Cross-Validation Estimate

<!-- badges: start -->

<!-- badges: end -->

We provide fast bootstrap methods to quantify the uncertainty of cross-validation estimate, which quickly estimates the standard error of the cross-validation estimate and produces valid confidence intervals for a population parameter measuring average model performance. The method overcomes the computational challenge inherent in bootstrapping the cross-validation estimate by estimating the variance component within a random effects model and is flexible for evaluating the performance of general predictive models. Details can be found in Cai et al. ([2023](https://doi.org/10.48550/arXiv.2307.00260)).

## Installation

You can install the released version of `BootCV` from [CRAN](https://cran.r-project.org/web/packages/available_packages_by_name.html#available-packages-G) with:

```{r install1}
# install.packages("BootCV")
```

And the development version from [GitHub](https://github.com) with:

```{r install2}
# install.packages("devtools")
# devtools::install_github("distant-spring/BootCV")
```

To use `BootCV`, we also need to install the required package `lme4` from [CRAN](https://cran.r-project.org/web/packages/lme4/index.html).

## Quick Start

This section gives users a general sense of the package with a basic example about using BootCV to generate cross-validation estimate and the corresponding confidence interval. We will briefly go over the key steps, basic operations, inputs and outputs.

First we load `BootCV` (and required package `lme4`)：

```{r}
# load required package
# library(lme4)
# library(BootCV)
```

We generate the data for illustration, which is a data matrix of dimension nobs x nvars:

```{r data}
set.seed(1)
# data generation
n <- 90
p <- 10
x <- matrix(rnorm(n * p), ncol = p)
beta <- rnorm(p)
y <- x %*% beta + rnorm(n)
data <- cbind(y, x)
```

Users need to specify the the summary statistics L for evaluation of model performance, which is calculated as a function of training data and testing data. Here we use the mean square error in linear regression as an example:

```{r L}
# summary statistics
L <- function(train.data, test.data) {
  y <- train.data[, 1]
  x <- train.data[, -1]
  yt <- test.data[, 1]
  xt <- test.data[, -1]

  fit <- lm(y ~ x)
  beta <- fit$coef

  return(mean((yt - cbind(1, xt) %*% beta)^2))
}
```

After specifying the sample size of training set in cross-validation, we can use `Boot.CV` to estimate the standard error of cross-validation estimate of summary statistics L:

```{r}
m <- 50 # training set size
# boot = Boot.CV(data, L, m)
# result with adjustment
# result1 = CV.confint(boot,data,L,m,method='Boot.CV',adj=T,print=T)
# result without adjustment
# result2 = CV.confint(boot,data,L,m,method='Boot.CV',adj=F,print=T)
```

`boot` is a list that contains all the relevant information for calculating the cross-validation estimate and confidence interval. Users can adjust the number of bootstraps, the number of cross-validations and the tuning parameter in determining the adjusted sample size of training set flexibly. Based on `boot`, users can use `CV.confint` to generate cross-validation estimation and confidence interval. Users can determine whether to adjust for the reduced training sample size and whether to print the results with the bool inputs `adj` and `print` and confidence level with `alpha`.

Sometimes, training the prediction model can be very expensive in terms of computation, and it may not be feasible to conduct many times bootstraps. Users may consider use `Boot.Cal` in this case. One thing to note is that we need to specify the confidence level when using `Boot.Cal` and keep it same as that used in `CV.confint` (default is 0.95).

```{r}
m <- 50 # training set size
# boot = Boot.Cal(data, L, m)
# result with adjustment
# result1 = CV.confint(boot,data,L,m,method='Boot.Cal',adj=T,print=T)
# result without adjustment
# result2 = CV.confint(boot,data,L,m,method='Boot.Cal',adj=F,print=T)
```

## More Examples

The algorithm quickly estimates the standard error of cross-validation estimate $$\widehat{Err}^{CV}_m=\frac{1}{B_{CV}}\sum_{b=1}^{B_{CV}} L\left\{D_{test}^b, \hat{\psi}(D_{train}^b)\right\},$$ where data $D=D_{train}^b\cup D_{test}^b$ represents the b-th split in cross-validation and summary statistics L evaluates the performance of estimation $\hat{\psi}$ fitted with training set in testing set. Many cross-validation applications can fit into this very general framework. This section provides several typical examples about how to construct the summary statistics L in different settings:

### Example 1 (Precision Medicine)

When the prediction model is a precision medicine strategy which is a binary classification rule to recommend a treatment to a patient based on his or her baseline characteristics to maximize the treatment benefit, we can construct an individualized treatment response (ITR) score by minimizing a loss function based on a training dataset $D_{train},$ $$\sum_{X_i\in D_{train}} \left\{Y_i-\gamma'\tilde{Z}_i-(G_i-\pi)\beta'\tilde{Z}_i \right\}^2 ,$$ where $X_i=(Z_i,G_i,Y_i),$ $Y_i$ is the response of interest with a higher value being desirable, $G_i \in \{0, 1\}$ is a binary treatment indicator and independent of the baseline covariate $Z_i$ (i.e., the treatment is randomly assigned to patients in the training set), $\tilde{Z}_i=(1, Z_i')'$ and $\pi=\Pr(G_i=1).$ Let the resulting minimizers of $\gamma$ and $\beta$ be $\hat{\gamma}(D_{train})$ and $\hat{\beta}(D_{train})$ (Tian et al., [2014](https://doi.org/10.1080/01621459.2014.951443)). The sample average of cross-validated treatment effect estimators is our final cross-validation estimator measuring the performance of the treatment recommendation system and the summary statistics measuring the prediction performance is:$$L\left(D_{test}, \psi\right)=\frac{\sum_{X_i\in D_{test}}Y_iG_iI(\psi'Z_i>0)}{\sum_{X_i\in D_{test}} G_i I(\psi'Z_i>0)}-\frac{\sum_{X_i\in D_{test}}Y_i(1-G_i)I(\psi'Z_i>0)}{\sum_{X_i\in D_{test}}(1-G_i)I(\psi'Z_i>0)}.$$ In this example, $X=(Z,G,Y)$ with $Z$, $G$ and $Y$ being predictors, the treatment assignment indicator, and a binary outcome, respectively, $\hat{\psi}(D_{train})=\hat{\beta}(D_{train})$.

### Example 2 (Binary Outcomes)

When the prediction model is constructed via fitting a logistic regression model, i.e., calculating a regression coefficient vector $\hat{\beta}(D_{train})$ by maximizing the log-likelihood function

$$ \sum_{(Z_i,Y_i)\in D_{train} } \left[ \beta'\tilde{Z}_i Y_i-\log\left\{1+\exp(\beta'\tilde{Z}_i)  \right\}\right],$$ based on a training dataset $D_{train},$ where $Z_i$ is the predictor, $\tilde{Z}_i=(1, Z_i')',$ and $Y_i\in \{0, 1\}$ is the binary outcome. If the dimension of $Z_i$ is high, a lasso-regularization can be used in estimating $\beta$. In any case, the c-index in a testing set $D_{test}$ can be calculated as $$\hat{\theta}(D_{train}, D_{test})=\frac{1}{\tilde{n}_{test, 0}\tilde{n}_{test, 1}}\sum_{X_i \in D_{test}(0)}\sum_{X_j\in D_{test1}(1)}I\left(\hat{\beta}(D_{train})'\tilde{Z}_i<\hat{\beta}(D_{train})'\tilde{Z}_j \right),$$ where $\tilde{n}_{test,g}$ is the number of observations in the set $D_{test}(g)=\{X_i=(Z_i,Y_i)\in D_{test}: Y_i=g\}, g\in \{0, 1\}.$ The sample average of cross-validated c-index estimators is our final cross-validation estimator measuring the classification performance of the logistic regression and the summary statistics measuring the prediction performance is:$$L\left(D_{test}, \psi\right)=\frac{1}{\tilde{n}_{test, 0}\tilde{n}_{test, 1}}\sum_{X_i\in D_{test}(0)}\sum_{X_j\in D_{test}(1)} I\left(\psi'\tilde{Z}_i<\psi'\tilde{Z}_j \right).$$ In this example, $X=(Z, Y)$ with $Z$ and $Y$ being the predictor and a binary outcome of interest, respectively, $\hat{\psi}(D_{train})=\hat{\beta}(D_{train})$, which can also be fitted with other methods such as random forest.

### Example 3 Mean Absolute Prediction Error (MAPE)

When the prediction model is constructed by fitting a standard linear regression model, i.e., calculating a regression coefficient vector $\hat{\beta}(D_{train})$ by minimizing a $L_2$ loss function $$ \sum_{X_i\in D_{train} } \left(Y_i- \beta'\widetilde{Z}_i \right)^2,$$ based on a training dataset $D_{train}$, where $Z_i$ is the baseline covariate for the $i$th patient and $\widetilde{Z}_i=(1, Z_i')'$ including an intercept. When the prediction error in testing set $D_{test}$ is calculated by mean absolute prediction error (Tian et al., [2007](https://doi.org/10.1093/biomet/asm036))$$\hat{\theta}(D_{train}, D_{test})=\frac{1}{n_{test}}\sum_{X_i \in D_{test}}|Y_i-\hat{\beta}(D_{train})'\tilde{Z}_i|,$$ where $n_{test}$ is the number of observations in the testing set. The sample average of cross-validated mean absolute prediction error estimators is our final estimator measuring the predictive performance of the linear model and the summary statistics measuring the prediction performance is:$$L\left(D_{test}, \psi\right)=\frac{1}{n_{test}}\sum_{X_i\in D_{test}}|Y_i-\psi'\widetilde{Z}_i|.$$ In this example, $X=(Z, Y)$ with $Z$ and $Y$ being the predictor and outcome of interest, respectively, $\hat{\psi}(D_{train})=\hat{\beta}(D_{train})$. If nonlinear prediction models are considered, then one may construct the prediction model via a more flexible machine learning algorithm such as random forest or neural network and the summary statistic is similar.

## References

Bryan Cai, Fabio Pellegrini, Menglan Pang, Carl De Moor, Changyu Shen, Vivek Charu and Lu Tian, 2023. "Bootstrapping the Cross-Validation Estimate." <https://doi.org/10.48550/arXiv.2307.00260>.

Lu Tian, Ash A Alizadeh, Andrew J Gentles and Robert Tibshirani. "A simple method for estimating interactions between a treatment and a large number of covariates." Journal of the American Statistical Association 109 1517--1532. <https://doi.org/10.1080/01621459.2014.951443>.

Lu Tian, Tianxi Cai, Els Goetghebeur, L. J. Wei. "Model evaluation based on the sampling distribution of estimated absolute prediction error." Biometrika, Volume 94, Issue 2, June 2007, Pages 297--311. <https://doi.org/10.1093/biomet/asm036>.
